---
title: "Homework #3"
author: "Natalie Mayer"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(lubridate)
library(stringr)

ca_rights <- read_csv(here("data", "california.csv"))
```


# Plot 1: Choropleth of water rights legal frameworks in the USA
The purpose of this visualization is to show that legal frameworks for water rights in the US are different for each state. This provides context for why Plot 2 is important. 

```{r}
library(sf)

shapefile <- read_sf(here("data", 
                          "shapefile", 
                          "s_18mr25.shp")) 

us_rights <- tibble(
  NAME = c("Arizona", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", 
            "Utah", "Wyoming", "Kansas", "North Dakota", "South Dakota", "Oregon", "Texas", 
            "Washington", "Nebraska", "Oklahoma", "New Hampshire", "Vermont", "Rhode Island", 
            "West Virginia", "Ohio", "Tennessee", "Missouri", "Louisiana", "Alabama", "Arkansas", 
            "Connecticut", "Delaware", "New York", "New Jersey", "Maryland", "Illinois", 
            "Indiana", "Iowa", "Kentucky", "Massachusetts", "Pennsylvania", "Mississippi", "Maine", "Michigan",
            "Minnesota", "North Carolina", "South Carolina", "Georgia", "Florida", 
            "Virginia", "Wisconsin", "California"),
  legal_framework = c(rep("Prior appropriation", 16), 
                        rep("Riparian", 31), 
                        "Combination"))

usa_states_joined <- left_join(shapefile, us_rights, by = "NAME") %>%
  drop_na(legal_framework) 
```

```{r}
legal_framework_pal <- c("Prior appropriation" = "#d8a451", 
                         "Riparian" = "#93afce", 
                         "Combination" = "#9b1b34") 

ggplot() + 
  geom_sf(data = usa_states_joined, 
          aes(fill = legal_framework), 
          color = "black", 
          size = 1) + 
  scale_fill_manual(values = legal_framework_pal) + 
  theme_void() + 
  labs(fill = "Legal Framework") +
  theme(legend.position = "bottom")
  
```





# Plot 2: CA Water Rights Uses by Decade
The purpose of this visualization is to show which water uses have priority over others. Due to the legal framework of water rights in CA (prior appropriation), older water rights are prioritized during times of water scarcity. 
- add California to title of plot

```{r}
water_use_pal <- c("Domestic" = "#9b1b34", 
                   "Environment" = "#93afce", 
                   "Industrial" = "#4a4343", 
                   "Agriculture" = "#d8a451")

ca_rights_use <- ca_rights %>%
  clean_names() %>%
  select(priority_date, cfs, water_use) %>%
  mutate(date = mdy(priority_date)) %>%
  mutate(date = if_else(year(date) > 2023, date - years(100), date)) %>%
  mutate(year = year(date))%>%
  filter(year > 1929) %>%
  filter(year < 2000) %>%
  mutate(decade = case_when(
    year %in% (1930:1939) ~ "1930s", 
    year %in% c(1940:1949) ~ "1940s", 
    year %in% (1950:1959) ~ "1950s", 
    year %in% c(1960:1969) ~ "1960s",
    year %in% (1970:1979) ~ "1970s", 
    year %in% c(1980:1989) ~ "1980s",
    year %in% c(1990:1999) ~ "1990s"
  )) %>%
  select(decade, cfs, water_use) %>%
  group_by(decade, water_use) %>%
  summarise(cfs = sum(cfs)) %>%
  ungroup() %>%
  filter(water_use != "other") %>%
  mutate(water_use = case_when(
    water_use %in% c("livestock", "irrigation") ~ "Agriculture",
    water_use %in% c("environmental", "fish") ~ "Environment",
    water_use == "industrial" ~ "Industrial",
    water_use == "domestic" ~ "Domestic")) %>%
  drop_na() %>%
  group_by(decade) %>%
  arrange(decade, desc(cfs)) %>%
  mutate(water_use = factor(water_use, levels = unique(water_use))) %>%
  ungroup()


ggplot(data = ca_rights_use, 
       aes(x = decade, y = cfs, fill = water_use)) + 
  geom_col(position = "dodge") + 
  scale_fill_manual(values = water_use_pal) + 
  scale_y_continuous(labels = scales::label_number(accuracy = 1, scale = 0.001, suffix = "K"),
                     expand = c(0,0)) +
  theme_classic() + 
  labs(subtitle = "Older water rights are more senior and are prioritized when water is scarce", 
       title = "Most California water rights are appropriated for industrial and domestic use",
       x = " ", 
       y = "Appropriated Water (CFS)", 
       fill = "Water Use",
       caption = "Data Source: HarDWR - Cumulative Water Rights Curves (Lisk et al. 2023)") + 
  theme(
    plot.title.position = "plot",
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(hjust = 0.5),
    
    legend.position = "bottom", 
    
    axis.text = element_text(color = "black")) + 
  coord_fixed(ratio = 0.00003)

```



# Plot 3: Relative change in appropriated water volume and annual runoff 
- need to add labels to each of the lines

```{r}
ca_rights_ts <- ca_rights %>%
  clean_names() %>%
  select(priority_date, cfs, water_use) %>%
  mutate(date = mdy(priority_date)) %>%
  mutate(date = if_else(year(date) > 2023, date - years(100), date)) %>%
  mutate(year = year(date)) %>%
  filter(year > 1929) %>%
  filter(year < 2000) %>%
  group_by(year, water_use) %>%
  summarise(cfs = sum(cfs)) %>%
  ungroup() %>%
  filter(water_use != "other") %>%
  mutate(water_use = case_when(
    water_use %in% c("livestock", "irrigation") ~ "Agriculture",
    water_use %in% c("environmental", "fish") ~ "Environment",
    water_use == "industrial" ~ "Industrial",
    water_use == "domestic" ~ "Domestic"
  )) %>%
  arrange(water_use, year) %>%  
  group_by(water_use) %>%     
  mutate(agg_cfs = cumsum(cfs)) %>%  
  ungroup()

```


```{r}
ggplot() + 
  geom_line(data = ca_rights_ts, 
            mapping = aes(x = year, y = agg_cfs, color = water_use), 
            linewidth = 1)+
  geom_hline(aes(yintercept = 36000), color = "lightblue", linetype = "dashed") + # max oroville dam cfs
  theme_classic() + 
  scale_y_continuous(expand = c(0,0),
                     labels = scales::label_number(accuracy = 1, scale = 0.001, suffix = "K")) +
  scale_x_continuous(breaks = seq(min(ag$year), max(ag$year), by = 10), 
                     expand = c(0,0)) + 
  coord_fixed(ratio = 0.00008) + 
  scale_color_manual(values = water_use_pal) +
  labs(y = "Appropriated Water (CFS)", 
       x = " ") + 
  annotate(
    geom = "text", 
    x = 1993, 
    y = 415000,
    label = "Industrial",
    size = 3.5,
    color = "#4a4343") + 
  annotate(
    geom = "text", 
    x = 1993, 
    y = 250000,
    label = "Domestic",
    size = 3.5,
    color = "#9b1b34") + 
  annotate(
    geom = "text", 
    x = 1985, 
    y = 192000,
    label = "Environment",
    size = 3.5,
    color = "#93afce") + 
  annotate(
    geom = "text", 
    x = 1992, 
    y = 158000,
    label = "Agriculture",
    size = 3.5,
    color = "#d8a451") + 
  annotate(
    geom = "text", 
    x = 1981, 
    y = 55000,
    label = "Maximum flow out of the Oroville Dam (36,000 CFS)",
    size = 2.5,
    color = "black") + 
  labs(title = "Growth of CA Appropriated Water Rights (1930-1999) by Beneficial Use",
       subtitle = "The Oroville Dam, the largest in the U.S. and the main reservoir of the State Water Project,\nserves as a reference for growing water appropriations, which have significantly increased\nparticularly for industrial use",
       caption = "Data Source: HarDWR - Cumulative Water Rights Curves (Lisk et al. 2023)",
       color = "Water Use") + 
  theme(plot.title.position = "plot", 
        plot.title = element_text(face = "bold"), 
        plot.caption = element_text(hjust = 0.5),
        legend.position = "bottom",
        
        axis.text = element_text(color = "black"), 
        axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0))
        )
```














