---
title: "Western U.S. Water Rights"
author: "Natalie Mayer"
format: html
editor: visual
---

# Data Wrangling

## Load data and libraries

```{r}
#| eval: true
#| warning: false
#| message: false
#| code-fold: true


library(tidyverse)
library(here)
library(stringr)
library(kableExtra)

arizona <- read_csv(here("data", "arizona.csv"))
california <- read_csv(here("data", "california.csv"))
colorado <- read_csv(here("data", "colorado.csv"))
idaho <- read_csv(here("data", "idaho.csv"))
montana <- read_csv(here("data", "montana.csv"))
nevada <- read_csv(here("data", "nevada.csv"))
new_mexico <- read_csv(here("data", "new_mexico.csv"))
oregon <- read_csv(here("data", "oregon.csv"))
utah <- read_csv(here("data", "utah.csv"))
washington <- read_csv(here("data", "washington.csv"))
wyoming <- read_csv(here("data", "wyoming.csv"))
```

## Reclassify variable types and add legal framework column

```{r}
#| eval: true
#| warning: false
#| message: false
#| code-fold: true


az <- arizona %>%
  mutate(legal_framework = "Prior Appropriation (But 40% of water comes from the Colorado River under a different system)") %>%
  mutate(waterRightID = factor(waterRightID),
         FIPS = factor(FIPS), 
         priorityDate = mdy(priorityDate),
         priorityDate = if_else(year(priorityDate) > 2023, priorityDate - years(100),
                                priorityDate),
         origWaterUse = factor(origWaterUse), 
         surBasinNum = factor(surBasinNum), 
         grdBasinNum = factor(grdBasinNum)) %>%
  filter(year(priorityDate) > 1925) %>%
  filter(year(priorityDate) < 2000)

ca <- california %>%
  mutate(legal_framework = "Hybrid: Riparian (Pre-1914) & Prior Appropriation (Post-1914)") %>%
  mutate(waterRightID = factor(waterRightID), 
         FIPS = factor(FIPS), 
         priorityDate = mdy(priorityDate),
         priorityDate = if_else(year(priorityDate) > 2023, priorityDate - years(100),
                                priorityDate), 
         origWaterUse = factor(origWaterUse), 
         basinNum = factor(basinNum)) %>%
  filter(year(priorityDate) > 1925) %>%
  filter(year(priorityDate) < 2000)

co <- colorado %>%
  mutate(legal_framework = "Prior Appropriation (Issued and adjudicated by Water Courts)") %>%
  mutate(waterRightID = factor(waterRightID),
         FIPS = factor(FIPS), 
         priorityDate = mdy(priorityDate),
         priorityDate = if_else(year(priorityDate) > 2023, priorityDate - years(100),
                                priorityDate),
         origWaterUse = factor(origWaterUse), 
         basinNum = factor(basinNum)) %>%
  filter(year(priorityDate) > 1925) %>%
  filter(year(priorityDate) < 2000)

id <- idaho %>%
  mutate(legal_framework = "Prior Appropriation and Basin Management Agencies") %>%
  mutate(waterRightID = factor(waterRightID), 
         FIPS = factor(FIPS), 
         priorityDate = mdy(priorityDate),
         priorityDate = if_else(year(priorityDate) > 2023, priorityDate - years(100),
                                priorityDate),
         origWaterUse = factor(origWaterUse), 
         basinNum = factor(basinNum)) %>%
  filter(year(priorityDate) > 1925) %>%
  filter(year(priorityDate) < 2000)

mt <- montana %>%
  mutate(legal_framework = "Prior Appropriation (Adjudicated by State Water Court)") %>%
  mutate(waterRightID = factor(waterRightID), 
         FIPS = factor(FIPS), 
         priorityDate = mdy(priorityDate),
         priorityDate = if_else(year(priorityDate) > 2023, priorityDate - years(100),
                                priorityDate),
         origWaterUse = factor(origWaterUse), 
         basinNum = factor(basinNum)) %>%
  filter(year(priorityDate) > 1925) %>%
  filter(year(priorityDate) < 2000)

nv <- nevada %>%
  mutate(legal_framework = "Prior Appropriation (Administered by State Engineer/Division of Water Rights)") %>%
  mutate(waterRightID = factor(waterRightID), 
         FIPS = factor(FIPS),
         priorityDate = mdy(priorityDate),
         priorityDate = if_else(year(priorityDate) > 2023, priorityDate - years(100),
                                priorityDate), 
         origWaterUse = factor(origWaterUse), 
         basinNum = factor(basinNum)) %>%
  filter(year(priorityDate) > 1925) %>%
  filter(year(priorityDate) < 2000)

nm <- new_mexico %>%
  mutate(legal_framework = "Prior Appropriation (Administered by State Engineer/Division of Water Rights)") %>%
  mutate(waterRightID = factor(waterRightID), 
         FIPS = factor(FIPS), 
         priorityDate = mdy(priorityDate),
         priorityDate = if_else(year(priorityDate) > 2023, priorityDate - years(100),
                                priorityDate), 
         origWaterUse = factor(origWaterUse), 
         basinNum = factor(basinNum)) %>%
  filter(year(priorityDate) > 1925) %>%
  filter(year(priorityDate) < 2000)

or <- oregon %>%
  mutate(legal_framework = "Prior Appropriation (Strict 5 year 'use-it-or-lose-it' window)") %>%
  mutate(waterRightID = factor(waterRightID), 
         FIPS = factor(FIPS), 
         priorityDate = mdy(priorityDate),
         priorityDate = if_else(year(priorityDate) > 2023, priorityDate - years(100),
                                priorityDate),
         origWaterUse = factor(origWaterUse), 
         basinNum = factor(basinNum)) %>%
  filter(year(priorityDate) > 1925) %>%
  filter(year(priorityDate) < 2000)

ut <- utah %>%
  mutate(legal_framework = "Prior Appropriation (Administered by State Engineer/Division of Water Rights)") %>%
  mutate(waterRightID = factor(waterRightID), 
         FIPS = factor(FIPS), 
         priorityDate = mdy(priorityDate),
         priorityDate = if_else(year(priorityDate) > 2023, priorityDate - years(100),
                                priorityDate),
         origWaterUse = factor(origWaterUse), 
         basinNum = factor(basinNum)) %>%
  filter(year(priorityDate) > 1925) %>%
  filter(year(priorityDate) < 2000)

wa <- washington %>%
  mutate(legal_framework = "Prior Appropriation and Water Resource Inventory Areas") %>%
  mutate(waterRightID = factor(waterRightID), 
         FIPS = factor(FIPS), 
         priorityDate = mdy(priorityDate),
         priorityDate = if_else(year(priorityDate) > 2023, priorityDate - years(100),
                                priorityDate), 
         origWaterUse = factor(origWaterUse), 
         basinNum = factor(basinNum)) %>%
  filter(year(priorityDate) > 1925) %>%
  filter(year(priorityDate) < 2000)

wy <- wyoming %>%
  mutate(legal_framework = "Prior Appropriation (Administered by State Engineer/Division of Water Rights)") %>%
  mutate(waterRightID = factor(waterRightID), 
         FIPS = factor(FIPS), 
         priorityDate = mdy(priorityDate),
         priorityDate = if_else(year(priorityDate) > 2023, priorityDate - years(100),
                                priorityDate),
         origWaterUse = factor(origWaterUse), 
         basinNum = factor(basinNum)) %>%
  filter(year(priorityDate) > 1925) %>%
  filter(year(priorityDate) < 2000)
```

## Join data for all western states

```{r}
#| eval: true
#| warning: false
#| message: false
#| code-fold: true


western_states <- bind_rows(az, ca, co, id, mt, nv, nm, or, ut, wa, wy) %>%
  mutate(waterUse = str_to_title(waterUse))
```

# Exploratory Data Analysis

## Q1: How have water rights changed over time in CA?

### Plot 1: Sankey Network

```{r}
#| eval: true
#| warning: false
#| message: false
#| code-fold: true


ca_wr <- ca %>%
  group_by(origWaterUse, waterUse) %>%
  summarise(cfs = sum(CFS)) %>%
  ungroup() %>%
  mutate(waterUse = str_to_title(waterUse))

library(networkD3)

# create links data frame
links <- data.frame(ca_wr) %>%
  mutate(source = origWaterUse, 
         target = waterUse, 
         value = cfs)

# create node data frame
nodes <- data.frame(
  name=c(as.character(links$source), 
  as.character(links$target)) %>% unique()
)

# reformat nodes so connection can be made using id
links$IDsource <- match(links$source, nodes$name)-1 
links$IDtarget <- match(links$target, nodes$name)-1

# make the network 
p <- sankeyNetwork(Links = links, Nodes = nodes,
              Source = "IDsource", Target = "IDtarget",
              Value = "value", NodeID = "name", 
              sinksRight=FALSE)
p

```

### Plot 2: Bar plots by water use

```{r}
ca_total <- ca %>%
  mutate(year = year(priorityDate)) %>%
  group_by(year, waterUse) %>%
  summarise(cfs = sum(CFS)) %>%
  ungroup() %>%
  filter(waterUse != "other")

ggplot(ca_total, aes(x = year, y = cfs, fill = waterUse)) + 
  geom_col() + 
  facet_wrap(~waterUse, scales = "free_y") + 
  theme_minimal()
```

```{r}
# Table of maximum year for water use 

data.frame(waterUse = c("Domestic", "Industrial", "Environmental", "Irrigation", "Livestock", "Fish"), 
       year = c("1927", "1977", "1977", "1927", "1981", "1972")) %>%
  kbl() %>%
  kable_classic()

```

-   The State Water Project began delivering water to So Cal in 1973. Interesting that the year prior saw an unprecedented surge in water rights to fish, and a few years later a maximum in water rights for industry, livestock, and environment.
-   1970 California Endangered Species Act, 1973 Federal Endangered Species Act

## Q2: How do Water Rights in CA compare to other Western States?

### Plot 3: Bar Plots

```{r}
#| eval: true
#| warning: false
#| message: false
#| code-fold: true

ws <- western_states %>%
  group_by(state, waterUse) %>%
  summarise(cfs = sum(CFS)) %>%
  ungroup() %>%
  drop_na()

ggplot(data = ws, aes(x = waterUse, y = cfs)) + 
  geom_col() + 
  coord_flip() +
  facet_wrap(~state, scales = "free_x") + 
  theme(axis.text.x = element_text(angle = 45)) 
```

### Plot 4: Geo Facet

```{r}
#| eval: true
#| warning: false
#| message: false
#| code-fold: true

az_use_prop <- az %>%
  group_by(waterUse, state) %>%
  summarise(cfs = sum(CFS)) %>%
  ungroup() %>%
  mutate(percent = round(cfs / sum(cfs) * 100, 0)) %>%
  select(waterUse, percent, state) %>%
  drop_na(waterUse) %>%
  filter(waterUse != "other")

ca_use_prop <- ca %>%
  group_by(waterUse, state) %>%
  summarise(cfs = sum(CFS)) %>%
  ungroup() %>%
  mutate(percent = round(cfs / sum(cfs) * 100, 0)) %>%
  select(waterUse, percent, state)

co_use_prop <- co %>%
  group_by(waterUse, state) %>%
  summarise(cfs = sum(CFS)) %>%
  ungroup() %>%
  mutate(percent = round(cfs / sum(cfs) * 100, 0)) %>%
  select(waterUse, percent, state)

id_use_prop <- id %>%
  group_by(waterUse, state) %>%
  summarise(cfs = sum(CFS)) %>%
  ungroup() %>%
  mutate(percent = round(cfs / sum(cfs) * 100, 0)) %>%
  select(waterUse, percent, state)

mt_use_prop <- mt %>%
  group_by(waterUse, state) %>%
  summarise(cfs = sum(CFS)) %>%
  ungroup() %>%
  mutate(percent = round(cfs / sum(cfs) * 100, 0)) %>%
  select(waterUse, percent, state)

nv_use_prop <- nv %>%
  group_by(waterUse, state) %>%
  summarise(cfs = sum(CFS)) %>%
  ungroup() %>%
  mutate(percent = round(cfs / sum(cfs) * 100, 0)) %>%
  select(waterUse, percent, state)

nm_use_prop <- nm %>%
  group_by(waterUse, state) %>%
  summarise(cfs = sum(CFS)) %>%
  ungroup() %>%
  mutate(percent = round(cfs / sum(cfs) * 100, 0)) %>%
  select(waterUse, percent, state)

or_use_prop <- or %>%
  group_by(waterUse, state) %>%
  summarise(cfs = sum(CFS)) %>%
  ungroup() %>%
  mutate(percent = round(cfs / sum(cfs) * 100, 0)) %>%
  select(waterUse, percent, state)

ut_use_prop <- ut %>%
  group_by(waterUse, state) %>%
  summarise(cfs = sum(CFS)) %>%
  ungroup() %>%
  mutate(percent = round(cfs / sum(cfs) * 100, 0)) %>%
  select(waterUse, percent, state)

wa_use_prop <- wa %>%
  group_by(waterUse, state) %>%
  summarise(cfs = sum(CFS)) %>%
  ungroup() %>%
  mutate(percent = round(cfs / sum(cfs) * 100, 0)) %>%
  select(waterUse, percent, state)

wy_use_prop <- wy %>%
  group_by(waterUse, state) %>%
  summarise(cfs = sum(CFS)) %>%
  ungroup() %>%
  mutate(percent = round(cfs / sum(cfs) * 100, 0)) %>%
  select(waterUse, percent, state)

state_water_prop <- bind_rows(
  az_use_prop, 
  ca_use_prop,
  co_use_prop,
  id_use_prop,
  mt_use_prop,
  nv_use_prop,
  nm_use_prop,
  or_use_prop,
  ut_use_prop,
  wa_use_prop,
  wy_use_prop
)
```

```{r}
library(geofacet)

custom_grid <- data.frame(us_state_grid1) %>%
  filter(name %in% c("Arizona", "California", "Colorado", "Idaho", 
                    "Montana", "Nevada", "New Mexico", "Oregon", 
                    "Utah", "Washington", "Wyoming"))

state_water_max <- state_water_prop %>%
  group_by(state) %>%
  filter(percent == max(percent)) %>%
  ungroup() %>%
  select(-c(percent)) %>%
  mutate(waterUse = str_to_title(waterUse))


wu_colors <- c("Other" = "#cad2d9", 
                 "Irrigation" = "#96ceb4", 
                 "Industrial" = "#ffcc5c", 
                 "Domestic" = "#bae1ff", 
                 "Fish" = "#ff6f69")


ggplot(state_water_max) + 
  geom_rect(aes(xmin = 0, xmax = 1, 
                ymin = 0, ymax = 1, 
            fill = waterUse)) + 
  scale_fill_manual(values = wu_colors) +
  geom_text(aes(x = 0.5, y = 0.7, label = state),
            family = "Comic Sans MS",
            face = "bold", 
            size = 3, 
            color = "black", 
            vjust = 2) + 
  facet_geo(~state, grid = custom_grid) + 
  coord_fixed(ratio = 1) +
  theme_void() + 
    theme(strip.text = element_blank()) + 
  labs(fill = "Water Use")
```
